---
title: "Lab on Functional Programming and R Notebooks"
output:
  pdf_document: default
  html_notebook: default
---

## R markdown

In this lab, we will run experiments using R notebook, which is a powerful programming- and note-taking tool that allows seamless integration between code and documentation. Here is an example:

```{r}
print("hello world!")
```

You may have seen Python notebook in TB1. R notebook is exactly the same idea, but designed for R programming language. Click the "green arrow" to run your code.

[[Markdown]{.underline}](https://en.wikipedia.org/wiki/Markdown) is a popular scripting language that allows you to write **formatted text** in *plain text editors* (such as notepad) and render it later. You can switch between the source code and visualization using the "source" and "visual" in the toolbar and see the differences.

## Targets for today

In today's lab, you will apply **functional programming** to code the pairwise distance function that we have encountered before many times. Specifically, given matrices $A \in \mathbb{R}^{n_1 \times m}$ and $B \in \mathbb{R}^{n_2 \times m}$, we would like to construct a matrix $D \in \mathbb{R}^{n_1 \times n_2}$, where $D_{i,j} = \mathrm{dist}(A_{[i,]}, B_{[j,]})$, $A_{[i,]}$ is $i$-th row of $A$ and $$\mathrm{dist}(a,b) = \sqrt{\sum_i (a_i - b_i)^2}.$$

## Preparation

Before you start, I strongly recommend you review [this week's lecture slides](https://github.com/anewgithubname/MATH10017-2022/blob/main/lecs/lec19.pdf) on list and functional programming.

**Write code below,** generating two random matrices $A$ and $B$ with 3 rows and 2 columns which are both filled with observations from standard normal distribution (Hint: `?rnorm`).

```{r}
#type your code here and run it by clicking the green arrow on the top-right corner.

A <- matrix(rnorm(3*2), 3,2)
B <- matrix(rnorm(3*2), 3,2)
```

Copy `pdist4` function from the previous lab (solution available online), and compute the distance matrix $D$ using $A$ and $B$. **NOTE**: in a previous lab we had $B \in \mathbb{R}^{m \times n_2}$ rather than $B \in \mathbb{R}^{n_2 \times m}$ as here. You will need to modify `pdist4` to take this into account.

```{r}
# Copy and modify the pdist4 code here
# fully vectorised version

# type your code here

```

## Pairwise Distance

We already know how to write vectorized code for computing $D$. Now let us inspect at this computation from a functional programming point of view.

-   Given two data matrices $A \in \mathbb{R}^{n_1 \times m}$ and $B\in \mathbb{R}^{n_2 \times m}$,
-   Fix $k$, and compute$(A_{[i,k]} - B_{[j,k]})^2$ for all $i, j$.
-   Store the outcome as the $i,j$-th element of a matrix $D_k$.
-   Apply the above computation for all $k$.
-   $D := \sqrt{\sum_k D_k}$

### Task 1

Write a function `diff2` that takes two scalars $a,b$, and compute $(a-b)^2$:

```{r}
#type your code here

```

`outer(X,Y,FUN)` applies function `FUN` two all pairs of elements in `X` and `Y`. In other words, it computes a matrix $D$, $D_{i,j} = \mathrm{fun}(X_i, Y_j), \forall i,j$. For example,

```{r}
a <- c(1,2,3,4)
b <- c(1,2,3,4)

a_plus_b <- function(a,b){
  return (a+b)
}

# function as an input variable
# outer applies the operation a_plus_b to each pair of elements in a and b
outer(a, b, a_plus_b)
# what is the i,j-th element in the matrix below?
```

Now use `outer` and `diff2`, compute a 3 by 3 matrix $D^{(1)}$, $D_{i,j}^{(1)} = \mathrm{diff2}(A_{[i,1]}, B_{[j,1]})$.

```{r}
#type your code here

```

### Task 2

We would like to repeat the above operation for all $k$. We can use a for loop but, in functional programming, we tend to think we "apply" a data operation to **the** **entire dataset** rather than create loops that iterate over each piece of our dataset.

This is exactly what we did above: We applied `diff2` to all pairs of $i,j$.

Let us create another function `D_k`, that takes one integer input $k$ and compute $D_k$ using `outer` (wrap the code you have just written in a function)

```{r}
# type your code here

```

Now apply this function to a list of numbers (hint: `?lapply`). The list contains integers from 1 to $m$, where $m$ is 2 in this toy case.

```{r}
# type your code here

```

You should get a list of matrices whose $k$ -th element is $D_k$.

### Task 3

Now, all that is left is summing $D_k$ together and we are done! Can we use a for loop to sum all $D_k$ matrices together?

In Functional Programming, we tend to think our program is a huge pipeline that applies functions repeatedly to our data: `op3(op2(op1(data)))...` (recall the "data pipeline" in the lecture slides)

We can rewrite the summation of $D_k$ using this paradigm:

$\sum_{k=1…4} D_k$ is the same as

`sum2(sum2(sum2(D_1,D_2),D_3),D_4)`, where `sum2(X,Y)` sums two matrices `X` and `Y` together. Write a function `sum2` that sums over two matrices

```{r}
# type your code here


```

Now we need to apply this function to the list of $D_k$ recursively. This can be done using a function called `Reduce`. `Reduce(f,x)` applies `f` to a list `x` recursively. In other words, it computes

`f(f(f(x_1, x_2),x_3),x_4)…`

```{r}
# type your code here

```

Square root the outcome.

```{r}
# type your code here

```

Check, is this the same result you obtained using `pdist4`?
```{r}
# Your checking code here

```

### Task 4

`lapply(l, fun)` applies a function `fun` to a list `l`.

There is a sibling function called `apply(A, MARGIN, fun)` which applies the function `fun` to a matrix along a certain margin. See `?apply`.

Setting `MARGIN` to 1 means you are applying `fun` to each row of the matrix.

Setting `MARGIN` to 2 means you are applying `func` to each column of the matrix.

`apply` applies `fun` to rows/columns of `A`, return a vector containing the results.

```{r}

X <- matrix(c(1,2,3,4), 2, 2)
X

# compute the average of a vector
average <-function(r) {
  return(sum(r)/length(r))
}

# compute row average
apply(X, 1, average)
```

Now, given the `D` matrix you obtained. Could you sort the elements of each row of `D`, in ascending order?

Hint: `sort` function, takes in a vector, and sort them according to the ascending order. `?sort`

```{r}
sort(c(4,3,2,1))
```

```{r}
# your code here 

```
